\documentclass[12pt]{article}

% --- Preamble ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{lmodern}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{amsmath,amsthm}
\let\lll\undefined % avoid conflict with babel
\usepackage{amssymb,mathtools,bm}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=teal}

\title{Lesson 01: Geometry of Linear Equations and Basics of Linear Algebra}
\author{}
\date{}

\begin{document}
\maketitle

\section*{Learning goals}
\begin{itemize}[label=$\triangleright$]
\item Understand the three “pictures” of a system $Ax=b$: column, row, and matrix view.
\item Distinguish between the cases: unique solution / infinitely many solutions / no solution.
\item Introduce: rank of a matrix, column space $\mathrm{col}(A)$, row space $\mathrm{row}(A)$, nullspace $\mathrm{null}(A)$.
\item First experiments with error, sensitivity, and least-squares solutions.
\end{itemize}

\section*{Quick reference}
\begin{itemize}[label=$\ast$]
\item \textbf{Column picture}: $Ax=b$ means $b$ is a linear combination of columns of $A$. If $b \in \mathrm{col}(A)$, the system is consistent.
\item \textbf{Row picture}: equations are dot products with rows of $A$, i.e. hyperplanes.
\item \textbf{Matrix picture}: $x \mapsto Ax$ is a linear operator transforming vectors.
\item $\mathrm{rank}(A)$ = number of independent columns = $\dim(\mathrm{col}(A))$.
\item $\mathrm{null}(A)=\{x:Ax=0\}$: family of solutions = one solution + any element of the nullspace.
\item If no solution exists $\Rightarrow$ use least squares (projection of $b$ onto $\mathrm{col}(A)$).
\end{itemize}

\section*{Exercises A – conceptual}
\begin{enumerate}[label=\textbf{A\arabic*}]
\item For $A=\begin{bmatrix}1&1\\1&1\end{bmatrix}$, for which $b=(b_1,b_2)^T$ does the system have a solution?
\item Find a nonzero vector in the nullspace of $A$ above. How many solutions does $Ax=0$ have?
\item If $A$ has size $m\times n$ and rank $r$, what is $\dim(\mathrm{null}(A))$?
\item For which relations between $m,n,r$ is the system typically: unique / infinite / LS?
\end{enumerate}

\section*{Exercises B – computational}
\begin{enumerate}[label=\textbf{B\arabic*}]
\item For $A=\begin{bmatrix}1&2&3\\2&4&6\\1&1&1\end{bmatrix}$:
\begin{enumerate}[label=(\alph*)]
\item compute $\mathrm{rank}(A)$,
\item give a basis for $\mathrm{col}(A)$,
\item describe $\mathrm{null}(A)$.
\end{enumerate}
\item Decide if $Ax=b$ with $b=(1,2,3)^T$ is consistent. If not – give a condition on $b$ for solvability.
\item Show that if $b \notin \mathrm{col}(A)$, the LS solution satisfies $A^TAx=A^Tb$.
\end{enumerate}

\section*{Exercises C – R laboratory}
\begin{enumerate}[label=\textbf{C\arabic*}]
\item \textbf{Three pictures}. In R, check whether $b$ is in $\mathrm{col}(A)$ by comparing $\mathrm{rank}(A)$ and $\mathrm{rank}([A|b])$.
\item \textbf{Rank and nullspace}. Use SVD to find a basis for the nullspace of a chosen matrix.
\item \textbf{Inconsistent system}. Construct $A,b$ with no exact solution. Compute the projection of $b$ onto $\mathrm{col}(A)$ and the LS solution.
\item \textbf{Sensitivity}. Generate an ill-conditioned matrix $A$, perturb $b$ with noise, and compare errors in $x$.
\item \textbf{Linear regression}. Compare LS solution via pseudoinverse with \texttt{lm()} in R.
\end{enumerate}

\section*{Mini-project (home assignment)}
\begin{itemize}
\item Generate $A$ ($m=40, n=3$) of full rank and a hidden $x^\ast$.
\item Add noise to $b$ with different levels $\sigma$.
\item For each $\sigma$: compute the LS solution, the error $\|x(\sigma)-x^\ast\|$, and save results to \texttt{.csv}.
\item Write a short report (3–5 sentences) about the impact of noise and $\kappa(A)$.
\end{itemize}

\section*{Discussion questions}
\begin{enumerate}[label=\textbf{D\arabic*}]
\item Why does the column picture immediately tell whether the system is consistent?
\item Does large rank always mean stability? What does $\kappa(A)$ really measure?
\item When do we prefer the exact solution (if it exists), and when LS?
\item Why is the LS residual $b-\hat b$ orthogonal to $\mathrm{col}(A)$?
\item Which of today’s phenomena will reappear with SVD/NMF/CUR on images?
\end{enumerate}

\section*{Suggested grading of homework}
\begin{itemize}
\item Correctness of calculations: 40\%.
\item Conclusions and interpretation: 40\%.
\item Repository/report cleanliness: 20\%.
\end{itemize}

\end{document}
